---
title: 'Week #9 Assignment'
author: "Poisson Regression"
date: "Due: Sunday, 11:59 PM"
output:
  html_document: 
    toc: no
    toc_float: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    fig_caption: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
    df_print: kable
  word_document: 
    toc: no
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: no
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 5
    fig_height: 4
---
<style type="text/css">
div#TOC li {
    list-style:none;
    background-image:none;
    background-repeat:none;
    background-position:0;
}
h1.title {
  font-size: 20px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 3 - and the author and data headers use this too  */
    font-size: 22px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}
h2 { /* Header 3 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 15px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}
</style>

```{r setup, include=FALSE}
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("openxlsx")) {
   install.packages("openxlsx")
   library(openxlsx)
}
#
knitr::opts_chunk$set(echo = TRUE,       
                      warning = FALSE,   
                      results = TRUE,   
                      message = FALSE
                      )  
```

\


\

\

* **Data Descriptions**

The daily total of bike counts was conducted monthly on the Brooklyn Bridge, Manhattan Bridge, Williamsburg Bridge, and Queensboro Bridge. To keep count of cyclists entering and leaving Queens, Manhattan, and Brooklyn via the East River Bridges. The Traffic Information Management System (TIMS) collects the count data. Each record represents the total number of cyclists per 24 hours at Brooklyn Bridge, Manhattan Bridge, Williamsburg Bridge, and Queensboro Bridge. 


* **Data Formats and Loading**

To save you time in finding a data set for Poisson regression, I created several subsets that contain the relevant information you need for this and next week's assignments. Here is the link to the Excel data set:[NYC cyclist data set (AssignDataSet.xlsx)](w09-AssignDataSet.xlsx). 

Please follow the indicated steps to find your data set:

```{r}
# You need to use library {openxlsx} to read the excel file
id=sample(1:10, 1)
dat <- read.xlsx("https://pengdsci.github.io/STA321/ww09/w09-AssignDataSet.xlsx", sheet = paste("data",id, sep = ""))
# I save the data in my download folder: C:\\Users\\75CPENG\\Downloads\\
# You should replace my local folder path 'C:\\Users\\75CPENG\\Downloads\\'
# with yours
write.csv(dat, paste("C:\\Users\\75CPENG\\Downloads\\",names(dat[6]), ".csv", sep=""))
# Go to your folder to load your data set (.csv) to your GitHub repository and 
# Read the data to your R directly to R so I can run your RMD source on my machine!
```

* **Assignment Instructions**

Your analysis and report should be similar to section 3 of my class note. PLEASE TELL STORIES BEHIND ALL R OUTPUTS (tables and figures) YOU GENERATED IN THE ANALYSIS. You can earn half of the credit if you only generate relevant outputs correctly but with no good storytelling. *The model diagnostic is not required in this assignment but will be required in the next assignment*.


The following components must be included in your analysis report.

* Your description of the data and the variables
  + data collection
  + variable names and definitions. Keep in mind that the variable **Date** is the observation ID.
  
* What is the research/practical question?
    + Explicit definition of the questions
    + what is the response variable that captures the information to address the research/practical questions
  
* What statistical model will be used in the analysis 
  + Assumptions and conditions - give a brief description of the models
  + Build a Poison regression model on the counts only. 
    - Use the p-values to perform variable selection - keep only significant variables in the final model
    - Interpret the regression coefficients of the Poison regression model
    - Explain/describe the steps of your analysis, motivation, and findings. 
  + Build a Poison regression on the proportions (rates) of cyclists entering and leaving the bridge in your data.
    - Use the p-values to select the significant variables
    - Interpret the regression coefficients of the Poison rate model
    - Explain/describe the steps of your analysis, motivation, and findings. 
* Summarize the findings of the above two models.
  
\

\







